```{r 02-data, echo=F}
#load(here('output','02-processed-data.rdata'))
```

# Quality control {#qc}

## Boxplot of sample raw intensities

Second, we want to inspect the distribution of peptide intensities between strains.
In addition, we will also observe in how many replicates each hit was quantified.

```{r boxplot-int_ub, echo=F, fig.cap='Distribution of expression for ubiquitous hits (i.e. detected in all strains)'}
# get lfq-peptide intensities (lfq=label-free quantitation) 
intensities = ms0 %>% dplyr::select(uniprot=majority_protein_i_ds,starts_with("lfq")) 

# Convert intensities to long format 
long_int_all = get_long_intensities(intensities)

# Get intensities for ubiquitously detected hits between strains
long_intensities = long_int_all %>% 
  filter(ratio_na_rep<1) %>% # At least one peptide in one replicate from a each strain
  group_by(strain) %>% mutate(nprot = n_distinct(uniprot)) %>% 
  hablar::convert(n_na_rep=int(n_na_rep)) %>% 
  arrange(n_na_rep)

# Get median expression and # proteins per strain as a function of NAs in replicates
stats_per_strain=long_intensities %>% 
  group_by(strain=toupper(strain),n_na_rep=factor(n_na_rep)) %>% 
  summarize(nprot=n_distinct(uniprot), MED=median_(log10_int)) %>% 
  mutate( YMAX = 12.5 - 0.25*as.numeric(n_na_rep))

# Compare intensities of ubiquitous hits between strains
bp=ggplot(data=long_intensities,mapping=aes(y=log10_int,x=toupper(strain))) +
  ggbeeswarm::geom_quasirandom(mapping = aes(group=n_na_rep,color=as.factor(nreplicates-n_na_rep)), na.rm=TRUE) +
  geom_violin(alpha=0.2, na.rm=TRUE,draw_quantiles = c(0.25,0.5,0.75)) + 
  geom_text(data=stats_per_strain,aes(label=nprot,color=nreplicates-n_na_rep,x=strain,y=YMAX),show.legend = F) +
  ylim(5,12.5) +
  ylab('Peptides intensities (log10)') + labs(x='Strains', colour='expressed in # replicates') +
  theme_pubclean() + grids(axis = 'y') + theme(axis.text = element_text(size=12)) +
  scale_color_canva(palette = "Pool party",limits=rev)
plot(bp)

BY_STRAINS = group_by(long_intensities,strain=toupper(strain)) %>% 
             summarize(median_exp=median_(log10_int))
BY_REP = group_by(long_intensities, n_replicates=as.integer(4-n_na_rep)) %>% 
  summarize(median_exp=median_(log10_int)) 

library(kableExtra)
TAB_STRAINS_EXP = kbl( transpose(BY_STRAINS) ,digits = 2) %>%
  kable_paper("striped", full_width = F) %>% kable_styling()

TAB_REP_EXP = kbl( transpose(round(BY_REP,1)) ) %>%
  kable_paper("striped", full_width = F) %>% kable_styling()  

```

**The distribution of peptide intensities between strains is relatively similar between strains:**
`r TAB_STRAINS_EXP`

**However, the peptide intensities can vary widely (1 unit = 10-fold) when a hit is not detected across all replicates:**
`r TAB_REP_EXP`

**On average, peptide intensities increase when a hit is found in more than one replicate**

We then look at hits quantified as a function of the number of strains in which 
at least one peptide was quantified. (non-ubiquitous hits)

```{r missing-hit, echo=F, fig.cap='Distribution of expression for hits not detected in all strains',out.width='100%'}
pep_missing = long_int_all %>%
  filter(ratio_na_rep==1) %>% 
  dplyr::select(-c(bio,tech)) %>% 
  distinct()

hit_missing = pep_missing$uniprot

bp_miss= long_int_all %>% arrange(n_na_strains) %>% 
         dplyr::filter(has_missing_strains) %>%
  ggplot(mapping=aes(y=log10_int,x=8-n_na_strains)) +
  geom_violin(mapping=aes(fill=factor(n_na_strains)),
              alpha=0.2, na.rm=TRUE,draw_quantiles = c(0.25,0.5,0.75),color='black',show.legend = F) +
  geom_text(aes(label=nprot_miss),y=12,check_overlap = T,col='red') + ylim(5,12) +
  ylab('Peptides intensities (log10)') + labs(x='expressed across # strains') +
  scale_x_continuous(breaks=0:8) +
  theme_pubclean() + grids(axis = 'y') + theme(axis.text = element_text(size=12))
  
plot(bp_miss)

stat_per_miss = long_int_all %>% 
  dplyr::select(n_na_strains,nprot) %>%
  mutate(n_strains= nstrains-n_na_strains) %>%
  distinct() %>% 
  ungroup() %>%
  arrange(desc(n_strains)) %>%
  mutate( f_exp = sprintf("%.1f%%",100*cumsum(nprot)/sum(nprot)) ) %>%
  dplyr::select(n_strains,n_na_strains,nprot,f_exp)


TAB_EXP_STRAIN = kbl(stat_per_miss,digits = 2) %>%
  kable_paper("striped", full_width = F) %>% kable_styling()

```

`r TAB_EXP_STRAIN`

**Similarly, we found a lower peptide intensity when a hit was found 
across less strains. Nevertheless, almost two thirds of the hits were detected in all strains.
Up to 87% of the hits were expressed across at least 5 out of 8 strains.**


## Number of missing values in samples

Furthermore, we can check whether the proportion of missing values is equally 
distributed among all samples.

```{r na-samples, echo=F,  out.width='100%', fig.cap='Number of quantified intensities across samples'}
# Count hits with NA in samples/groups
na_int= intensities %>% 
        group_by(uniprot) %>%
        summarize(across(all_of(paste0("lfq_intensity_",samples)), is.na ) )
colnames(na_int) = c('uniprot',samples)

ComplexUpset::upset(data=as.data.frame(1-na_int[,-1]),intersect=samples,
      name='intensity quantified', min_degree=5, sort_sets=F, 
      width_ratio=0.1,min_size=10, keep_empty_groups=F, wrap=TRUE,
         base_annotations=list(
          'Number of hits'=ComplexUpset::intersection_size(text_colors=c(on_background='red', on_bar='yellow'),text=list(angle=0,hjust=0.45,size=2.8))
           + annotate( geom='text', x=Inf, y=Inf, label=paste('Total hits:', nrow(na_int)), vjust=1, hjust=1 )
         ),
)


```


## Compare all-vs-all expression

### Scatterplots

```{r scatterplot-all, fig.path='plot/', fig.show='hold', warning=F, out.width='100%'}
# scatterplots between all samples
scmat_all=draw_scatterplots(datain=int_all)
print(scmat_all)

```

### Heatmap correlation 

```{r heatmap-all, fig.path='plot/', fig.show='hold', warning=F, out.width='100%' }
# heatmap correlation
cs_all=compute_samples_correlation(int_all)
COR_RANGE = range( cs_all[row(cs_all) == (col(cs_all) - 1)] )

by_sample = df.group %>% column_to_rownames("sample")
hm_all=draw_heatmap_samples(mcor = cs_all,df.group=by_sample,col.group = col.group)
```


The heatmap correlations of all samples show the high correlation of expression 
between replicates and between most strains:
`r sprintf("[%.3f - %.3f]",COR_RANGE[1],COR_RANGE[2])`

The expression from the 1st biological replicates of strains CQC and CMP seem
slighlty less correlated to the other samples.

### Principle Component Analysis

Finally, the PCA  reveals the distance between each sample.

```{r pca-all, fig.path='plot/', fig.show='hold', warning=F, out.width='100%'}
int_all_scaled = scale(int_all,center=T, scale = T)
make_pca(na.omit(int_all_scaled), with_labels=T,col_by_group=1:4)
```


```{r 03-data, echo=F}
save.image(here('output','03-quality-control.rdata'))
```