# Strains expression comparison

## Scatterplots 

Initially, we can look at the scatterplots of intensities all-versus-all samples.

```{r scatterplot-strains, fig.path='plot/', fig.show='hold', out.width='100%',warning=F}
# scatterplots between strains
scmat_strains=draw_scatterplots(datain=int_filt_strains)
print(scmat_strains)
```

## Heatmap Correlation

We then compute spearman rank correlations of intensities between all samples.

```{r heatmap-strains}
# heatmap correlation
cs_strains=compute_samples_correlation(int_filt_strains)
COR_RANGE = range( cs_strains[row(cs_strains) == (col(cs_strains) - 1)] )

hm_strains=draw_heatmap_samples(mcor = cs_strains,df.group=c(),col.group = col.group)
```

The heatmap correlations show the high correlation of expression between strains:
`r sprintf("[%.3f - %.3f]",COR_RANGE[1],COR_RANGE[2])`


## Principle Component Analysis

Finally, the PCA  reveals the distance between each sample.

```{r pca-strains}
int_scaled_strains = scale(int_filt_strains,center=T, scale = T)
make_pca(na.omit(int_scaled_strains), with_labels=F,col_by_group=1:2)
```
