```{r load-data-comparison, echo=F}
chap_cur=06
#load(here::here('output',sprintf("%02d-strain-comparison.rdata",chap_cur-1)))
```

# Differential Expression

We then compare the protein expression between strains for each pairwise
comparison (28 unique pairwise combinations)

To detect differentially expressed genes, we use the limma analysis on
normalized protein expression.

## Volcano plots

First, we can look at the volcano plots (log-foldchange vs qvalue) for each unique
pairwise comparison.

```{r volcano-plots, fig.path='plot/', fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}
#ind_na_rows  = find_na_rows(int_norm,as.indices = T)
df_imputed = tibble( uniprot = int_norm$uniprot,
                     is_imputed = (rowSums( is.na(int_norm))>0)* 1, 
                     imputed = factor(is_imputed, levels = c(0,1), labels = c("not", "is_imputed")))
#int_norm_bpca$imputed = rowSums( is.na(int_norm))>0
volcPlot(INPUT=int_norm_bpca, IMPUTED=df_imputed,MIN_LFC=2, MIN_PVAL=0.01, WHICH='both', TOPN = 20)

```

## Differentially expressed genes

```{r dfe-gene }
# Without NA
volcano_data =  get_volcano_data(input_data=int_norm_nona, which='both',
                            min_lfc=2, min_pval=0.01, topn = 20)
df_limma = bind_rows(volcano_data) %>% as_tibble()
dfe_nona = subset(df_limma,sig!='Non significant')
N_DFE_NONA = nrow(dfe_nona)
NPROT_DFE_NONA =  n_distinct(dfe_nona$ID)

# With imputed expression to replace NA
dfe =  get_volcano_data(input_data=int_norm_bpca, which='both',
                            min_lfc=2, min_pval=0.01, topn = 20) %>% 
                bind_rows %>% as_tibble() %>% filter(sig!='Non significant')

N_DFE_IMPUTE = nrow(dfe)
NPROT_DFE_IMPUTE =  n_distinct(dfe$ID)

down = dfe %>% group_by(ID) %>% dplyr::filter(sig=='Downregulated') %>%
       summarize( strains_down = paste0(comparison,collapse=' '),
                  down_AMH = str_count(strains_down,'AMH-'),
                  down_BAN = str_count(strains_down,'BAN-'),
                  down_BED = str_count(strains_down,'BED-'),
                  down_BPL = str_count(strains_down,'BPL-'),
                  down_BTT = str_count(strains_down,'BTT-'),
                  down_CMP = str_count(strains_down,'CMP-'),
                  down_CPI = str_count(strains_down,'CPI-'),
                  down_CQC = str_count(strains_down,'CQC-'))
up = dfe %>% group_by(ID) %>% dplyr::filter(sig=='Upregulated') %>%  
      summarize( strains_up = paste0(comparison,collapse=' '),
                  up_AMH = str_count(strains_up,'AMH-'),
                  up_BAN = str_count(strains_up,'BAN-'),
                  up_BED = str_count(strains_up,'BED-'),
                  up_BPL = str_count(strains_up,'BPL-'),
                  up_BTT = str_count(strains_up,'BTT-'),
                  up_CMP = str_count(strains_up,'CMP-'),
                  up_CPI = str_count(strains_up,'CPI-'),
                  up_CQC = str_count(strains_up,'CQC-'))
# get_dfe(int_norm, MIN_LFC=2, MIN_PVAL=0.01,  WHICH='both', TOPN = 20) %>% remove_rownames() %>% 
#   dplyr::left_join(sc_identifiers, by=c('ID'='UNIPROT'))

# Number of times a hit is differentially expressed
df_dfe = dfe %>% left_join(janitor::tabyl(dfe,ID,sig)) %>%
         left_join(down) %>% left_join(up) %>%
         rename(uniprot=ID) %>% 
         group_by(uniprot,comparison) %>% 
        mutate( n_strains_up = sum(c_across(starts_with('up_')) !=0 ),
                n_strains_down = sum(c_across(starts_with('down_'))!=0)) %>%
        replace_na(list(n_strains_up=0,n_strains_down=0)) %>%
        left_join(evo_yeast, by=c('uniprot'='UNIPROT')) 
  

df_dfe_annot = df_dfe %>%
          left_join(sc_annotation_orf,by=c('uniprot'='UNIPROT')) %>%
  mutate(uniprot_link = paste0("<a href='https://www.uniprot.org/uniprot/",uniprot,"'>",uniprot,"</a>"),
         sgd_link = paste0("<a href='https://www.yeastgenome.org/locus/",SGD,"'>",SGD,"</a>"),
         regulated = Downregulated+Upregulated) %>% 
  dplyr::relocate(uniprot,uniprot_link,sgd_link,regulated,Downregulated,Upregulated, 
                  GENENAME,ORF,PNAME,'FUNCTION','BIOPROCESS_all','ORTHO','OTHER')
```

We get **`r NPROT_DFE_NONA`** genes differentially expressed when
excluding genes with missing expression in any samples. 

On average, each gene is detected in `r N_DFE_NONA/NPROT_DFE_NONA` unique pairwise strain comparison.


After imputation of missing expression with bpca (Bayesian missing value imputation),
we get **`r NPROT_DFE_IMPUTE-NPROT_DFE_NONA`** more genes differentially expressed (n=**`r NPROT_DFE_IMPUTE`**).

On average, each gene is detected in `r N_DFE_IMPUTE/NPROT_DFE_IMPUTE` unique pairwise strain comparison.


The following table shows the list of differentially expressed genes across all unique pairwise comparison,
with annotations data and conservation/snp information.

```{r dfe-table,fig.path='plot/', fig.show='hold', warning=F, lightbox=T, gallery=T}
library(kableExtra)
library(formattable)
library(DT)

ft_dt = df_dfe_annot %>% 
  formattable(
    list(
      `Downregulated` = color_tile("white", "red"),
      `Upregulated` = color_tile("white", "blue"),
      `regulated` = color_tile("white", "gray")
    )
) %>% as.datatable(
        options = list(
            fixedHeader=T,
            paging = TRUE, pageLength = 20,  ## paginate the output and #rows for each page
            scrollY = TRUE,   ## enable scrolling on X/Y axis
            #autoWidth = TRUE, ## use smart column width handling
            server = FALSE,   ## use client-side processing
            dom = 'Bfrtip', buttons = list('csv', 'excel', list(extend = 'colvis')),
            fixedColumns = list(leftColumns = 1),
            columnDefs = list(list(width = '50px', visible=TRUE, targets = "_all"))
          ),
  extensions = c('FixedHeader','FixedColumns','Buttons'),
  selection = 'single',           ## enable selection of a single row
  filter = 'bottom',              ## include column filters at the bottom
  rownames = FALSE,               ## don't show row numbers/names
  width = NULL, 
  height = NULL,
  caption = NULL
) %>% 
   formatStyle(columns = 1:30, target= 'row',lineHeight='100%', `font-size` = '12px')

ft_dt

```


## Heatmap of expression for differentially expressed genes
```{r heatmap-exp, fig.height=25, fig.ext=c('pdf','png'), fig.path='plot/', fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}

dat_scaled = int_scaled_strains %>% as.data.frame() %>% rownames_to_column('uniprot') 

dfe_exp = dat_scaled %>% dplyr::filter( uniprot %in% dfe$ID) %>% 
          left_join(sc_identifiers,by=c('uniprot'='UNIPROT'))  %>% 
          dplyr::filter(!duplicated(GENENAME)) %>%
          mutate(GENENAME=if_na(GENENAME,uniprot))%>%
          column_to_rownames(var = 'GENENAME') %>%
          dplyr::select(-ORF,-uniprot,-SGD)

p_dfe_exp=pheatmap::pheatmap(dfe_exp,
                             fontsize = 5,cellwidth = 5,cellheight =5,border_color = NA,treeheight_col = 10,
                             filename = here('plot','heatmap-exp.pdf'))
knitr::include_graphics(here('plot','heatmap-exp.pdf'))
```


## Heatmap of expression differences

```{r heatmap-diffexp, fig.height=25, fig.path='plot/',  fig.ext=c('pdf','png'), fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}


dfe_lfc = get_volcano_data(input_data=int_norm_bpca, which='both',min_lfc=2, min_pval=0.01, topn = 20) %>% 
          bind_rows %>% as_tibble() %>%
          pivot_wider(id_cols=ID, names_from = 'comparison', values_from = 'EffectSize') %>% 
          dplyr::filter(ID %in% dfe$ID) %>% 
          left_join(sc_identifiers,by=c('ID'='UNIPROT')) %>%
          dplyr::filter(!duplicated(GENENAME)) %>%
          mutate(GENENAME=if_na(GENENAME,ID))%>%
          column_to_rownames('GENENAME')

dfe_lfc_mat = dfe_lfc %>% dplyr::select(-ORF,-ID,-SGD)
# Heatmap of differentially expressed genes
pheatmap::pheatmap(dfe_lfc_mat,fontsize = 5,cutree_rows = 10,cellwidth = 5,cellheight =5,border_color = NA,
                          treeheight_col = 10, filename = here('plot','heatmap-lfc.pdf')  )
knitr::include_graphics(here('plot','heatmap-lfc.pdf'))
```

## Cluster genes based on their profile of differential expression strain-pairwise 

```{r clust-prot, fig.height=20, fig.path='plot/',  fig.ext=c('pdf','png'), fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}
# Transpose the matrix to calculate distance between experiments row-wise
d_strain <- dfe_lfc_mat %>% t() %>% dist(.,method = "euclidean", diag = FALSE, upper = FALSE)
# Calculate the distance between proteins row-wise 
d_prot <- dfe_lfc_mat %>% dist(.,method = "euclidean", diag = FALSE, upper= FALSE)

hc_prot = hclust(d_prot,method='ward.D2')
cl_prot = hc_prot %>% cutree(k=10)
library(dendextend)
dend_prot = as.dendrogram(hc_prot)
dend_prot = rotate(dend_prot,seq_along(cl_prot))
dend_prot <- color_branches(dend_prot, k=10)
dend_prot <- color_labels(dend_prot, k=10)

#labels_colors(dend) <-rainbow_hcl(3)[sort_levels_values( as.numeric(dend_prot[,5])[order.dendrogram(dend)] )]
dend_prot <- hang.dendrogram(dend_prot,hang_height=0.1)
dend_prot <- set(dend_prot, "labels_cex", 0.5)
# And plot:
par(mar = c(2,3,0,3))
plot(dend_prot,horiz =  TRUE,  nodePar = list(cex = .007))

#legend("topleft", legend = iris_species, fill = rainbow_hcl(10))
par(mar = rep(0,4))
circlize_dendrogram(dend_prot)

```


## Functional map for differentially expressed genes

```{r treemap-bioprocess, fig.path='plot/', fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}
library(treemap)
library(d3Tree)
treemap(df_dfe_annot, index=c("BIOPROCESS_all", "comparison"), vSize='regulated', vColor="EffectSize", type="value") 
treemap(df_dfe_annot, index=c("BIOPROCESS_all", "GENENAME"), vSize="regulated", vColor="EffectSize", type="value",)
```

## Functional enrichment for differentially expressed genes

```{r func-enrich, fig.height=20, fig.path='plot/',  fig.ext=c('pdf','png'), fig.show='hold', warning=F, lightbox=T, gallery=T, out.width='100%'}
library(enrichplot)
library(clusterProfiler)

all_genes = int_scaled_strains %>%
            as.data.frame() %>%
            rownames_to_column('ID') %>% 
            left_join(sc_identifiers,by=c('ID'='UNIPROT')) %>%
            dplyr::filter(!duplicated(GENENAME)) %>%
            mutate(GENENAME=if_na(GENENAME,ID)) %>% 
            as_tibble() %>% 
            dplyr::select(where(is.character))


genes_proteomics = all_genes$ORF
genes_dfe = dfe_lfc$ORF
genes_dfe_uni = dfe_lfc$ID

genes_dfe_list = df_dfe_annot %>% group_by(ORF) %>% summarize( mean_lfc = mean(EffectSize)) %>% arrange(desc(mean_lfc)) %>% pull(mean_lfc,ORF) 
genes_dfe_list_uni = df_dfe_annot %>% group_by(uniprot) %>% summarize( mean_lfc = mean(EffectSize)) %>% arrange(desc(mean_lfc)) %>% pull(mean_lfc,uniprot)


# ggo <- groupGO(gene     = gene,
#                OrgDb    = org.Sc.sgd.db,
#                ont      = "CC",
#                level    = 3,
#                readable = TRUE)
library(org.Sc.sgd.db)
# ego <- enrichGO(gene          = genes_dfe,
#                 keyType       = "ORF",
#                 universe      = genes_proteomics,
#                 OrgDb         = org.Sc.sgd.db,
#                 ont           = "ALL",
#                 pAdjustMethod = "BH",
#                 pvalueCutoff  = 0.01,
#                 qvalueCutoff  = 0.05,
#                 readable      = F,
#                 pool=T)
# dotplot(ego)

ego3 <- gseGO(geneList     = genes_dfe_list,
              OrgDb        = org.Sc.sgd.db,
              keyType       = "ORF",
              ont          = "ALL",
              minGSSize    = 20,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
dotplot(ego3)
heatplot(ego3,foldChange = genes_dfe_list)

#kk <- enrichKEGG(gene         = gene,organism     = 'sce', pvalueCutoff = 0.05)
kk2 <- gseKEGG(geneList     = genes_dfe_list_uni,organism     = 'sce', minGSSize    = 20,keyType = 'uniprot',
               pvalueCutoff = 0.05,verbose      = FALSE)
heatplot(kk2, foldChange=genes_dfe_list_uni)

mkk <- enrichMKEGG(gene = genes_dfe_uni,
                    organism = 'sce',
                    pvalueCutoff = 1,
                    qvalueCutoff = 1,
                    keyType = 'uniprot')
dotplot(mkk)

```

## Comparison of functional enrichment for clusters of differentially expressed genes


```{r func-enrich-clust}
genes_dfe_cluster = split(dfe_lfc$ORF,cl_prot)

ck <- compareCluster(geneCluster = genes_dfe_cluster, fun = enrichGO,
                     keyType= "ORF",
                     ont='ALL',
                     universe=genes_proteomics,
                     OrgDb= org.Sc.sgd.db,
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.01,
                     qvalueCutoff  = 0.05,pool=F)
dotplot(ck)
```

```{r save-data-diffexp, echo=F}
save.image(here('output',sprintf("%02d-differential-expression.rdata",chap_cur)))
```