# Session Info {.tabset}

## Platform info 
```{r platform, out.width='100%',}
df_platform = devtools::session_info()$platform %>% 
  unlist(.) %>% 
  as.data.frame(.) %>% 
  rownames_to_column(.)

kable(
  df_platform,
  booktabs = T, 
  align = "l",
  caption = "(Reproducibility - platform info)", # complete caption for main document
  caption.short = " " # 
)

```

## Packages loaded and used (i.e. attached to the workspace)

```{r loaded-packages, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", out.width='100%', results='asis'}
library(sessioninfo)
df_session_packages <- devtools::session_info()$packages %>% 
  as.data.frame(.) %>% 
  filter(attached == TRUE) %>% 
  dplyr::select(loadedversion, date) %>% 
  rownames_to_column
colnames(df_session_packages) <- c("Package", "Loaded version", "Date")

kable(
  df_session_packages, 
  booktabs = T, 
  align = "l",
  caption = "(Reproducibility - loaded packages)", # complete caption for main document
  caption.short = " " # 
) %>% 
  kable_styling(full_width = F,
                latex_options = c(
                  "hold_position" # stop table floating
                ) 
  ) 

pkgs=capture.output(dput(df_session_packages$Package)) %>% paste(collapse='')

cmd_xfun_install="install.packages('xfun')\n"
cmd_pkg_load=sprintf("xfun::pkg_load2(%s)",pkgs)

```

To reproduce this Rmarkdown document, you may need to run the following lines
for loading/installing the necessary packages:

Copy paste the code below to your rsession:

`r cmd_xfun_install`

`r cmd_pkg_load`

Pay attention to the error message you might get.

Common errors could be due to:

* Not having admin privileges to write in the home R directory
* Some package installation requires prior linux libraries to be installed
* Some packages might not be available for older R version
* Some packages are not obtained from CRAN but from Bioconductor (it may require to use the function BiocManager::install())

## All packages required (i.e. including dependencies)

```{r all-packages, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", out.width='100%', results='asis'}
df_all_packages= devtools::session_info()$packages %>% as.data.frame() %>% pull(package)

kable(devtools::session_info()$packages,
      row.names = T,
      caption='all packages used (include dependencies)')

```

```{r session-info}
sessioninfo::session_info()
```